Containers and R---the Rockerverse and beyond

### 2019-11-06 manuscript.Rmd

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2
    - 48149 Münster
    - Germany
    - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
        - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263 USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263 USA
    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
---

@@

- Binder
  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
- GPU [@noamross, @cboettig]
- Gigantum stack

>>>>

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2
    - 48149 Münster
    - Germany
    - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
        - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science, University of California, Berkeley, CA 94720 USA
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263 USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263 USA
    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
---

@@

- Binder

The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others. A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within. From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.

To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository. Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within. While this approach works well for most run of the mill Python projects, it is not so seamless for R projects. For any R projects that use the Tidyverse suite [citation in review. Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.

[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users. `holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images. Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute. `holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [doi:10.1080/00031305.2017.1375986]. The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.

  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
- GPU [@noamross, @cboettig]
- Gigantum stack
